Bayesian Analysis of Gas-free Galaxies in Simulations I: Reviving Stochasticity

### 2023-03-01 papers/paper-1/sections/appendix.tex
Small updates in preparation for SPZ meeting
Start adding small section on hardening rate

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_gqs_prior}.pdf}
    \caption{\drafting{Prior predictive distributions of latent parameters. Note the mode of the distributions roughly coincide with what one would expect from the literature.}}
    \label{fig:prior_latent}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_prior_pred_log10_proj_density_mean}.pdf}
    \caption{\drafting{Prior predictive estimate for the core-S\'ersic fit. Darker shaded regions correspond to higher Bayesian Credible Intervals (BCIs). The observed data is overlaid. From the prior predictive estimate, it is seen that the model and the assumed parameter hyperdistributions are able to describe the observed data, in that the observed data lies within the range allowed by the prior predictive estimate.}}
    \label{fig:priorpred}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_pair_0}.pdf}
    \caption{\drafting{Corner plot of hierarchical model hyperparameters. Each of the hyperdistributions are modelled as Gamma distributions with shape parameters $\alpha$ and $\beta$. For example, the hyperdistribution on the core radius $r_\mathrm{b}$ is modelled as $r_\mathrm{b} \sim \text{Gamma}(r_{\mathrm{b},\alpha}, r_{\mathrm{b},\beta})$. Critically, the hyperdistributions are unimodal with highest-density regions located some distance from the distribution edges, indicating the HMC sampling has converged well. The contours correspond to the 50\%, 90\%, 95\%, and 99\% BCIs. }}
    \label{fig:corner_params_1}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_pair_1}.pdf}
    \caption{\drafting{Same as \autoref{fig:corner_params_1}}}
    \label{fig:corner_params_2}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_pair_2}.pdf}
    \caption{\drafting{Same as \autoref{fig:corner_params_1}}}
    \label{fig:corner_params_3}
\end{figure*}

>>>>

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_gqs_prior}.pdf}
    \caption{\drafting{Prior predictive distributions of latent parameters. Note the mode of the distributions roughly coincide with what one would expect from the literature.}}
    \label{fig:prior_latent}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_prior_pred_log10_proj_density_mean}.pdf}
    \caption{\drafting{Prior predictive estimate for the core-S\'ersic fit. Darker shaded regions correspond to greater Highest Density Intervals (HDIs). The observed data is overlaid. From the prior predictive estimate, it is seen that the model and the assumed parameter hyperdistributions are able to describe the observed data, in that the observed data lies within the range allowed by the prior predictive estimate.}}
    \label{fig:priorpred}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_pair_0}.pdf}
    \caption{\drafting{Corner plot of hierarchical model hyperparameters. Each of the hyperdistributions are modelled as Gamma distributions with shape parameters $\alpha$ and $\beta$. For example, the hyperdistribution on the core radius $r_\mathrm{b}$ is modelled as $r_\mathrm{b} \sim \text{Gamma}(r_{\mathrm{b},\alpha}, r_{\mathrm{b},\beta})$. Critically, the hyperdistributions are unimodal with highest-density regions located some distance from the distribution edges, indicating the HMC sampling has converged well. The contours correspond to the 50\%, 90\%, 95\%, and 99\% HDIs. }}
    \label{fig:corner_params_1}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_pair_1}.pdf}
    \caption{\drafting{Same as \autoref{fig:corner_params_1}}}
    \label{fig:corner_params_2}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_pair_2}.pdf}
    \caption{\drafting{Same as \autoref{fig:corner_params_1}}}
    \label{fig:corner_params_3}
\end{figure*}

### 2023-03-01 papers/paper-1/sections/results.tex
Start adding small section on hardening rate

\drafting{ENSURE CONSISTENCY WITH BCI AND HDI. SAME THING, JUST DIFFERENT NAME}

@@

\begin{table}
    \centering
    \caption{The parameters for each hyperparameter distribution for the core-S\'ersic model, where each distribution is a gamma distribution.}
    \label{tab:cshypers}
    \begin{tabular}{ccc}
        \hline
        Hyperparameter & $\mu$ & $\sigma$ \\
        \hline
        $r_{\mathrm{b}, \mu}$ & 0.0 & 2.0 \\
        $r_{\mathrm{b}, \sigma}$ & 0.0 & 1.0 \\
        $R_{\mathrm{e}, \mu}$ & 0.0 & 14.0 \\
        $R_{\mathrm{e}, \sigma}$ & 0.0 & 6.0 \\
        $\tilde{\Sigma}_{\mathrm{b}, \mu}$ & 0.0 & 20.0 \\
        $\tilde{\Sigma}_{\mathrm{b}, \sigma}$ & 0.0 & 10.0 \\
        $\gamma_\mu$ & 0.0 & 0.5 \\
        $\gamma_\sigma$ & 0.0 & 1.0 \\
        $n_\mu$ & 0.0 & 8.0 \\
        $n_\sigma$ & 0.0 & 5.0 \\
        $a_\mu$ & 0.0 & 20.0 \\
        $a_\sigma$ & 0.0 & 10.0 \\
        \hline
    \end{tabular}
\end{table}

@@

\subsubsection{Observable Quantities}
The hierarchical model takes three observable quantities, denoted by hat-variables in red circles in \autoref{fig:graphicalmodel}. 
These variables are the radius $\hat{R}$, and the mean projected mass density $\hat{\Sigma}$, and the standard deviation of the projected mass density $\hat{\sigma}_\Sigma$ measured in the radial bins centred on $\hat{R}$.
For each \child{} simulation, we choose to `observe' the first snapshot where the binary semimajor axis is \textit{at most} $10\,\mathrm{pc}$. 
In the event that there is no snapshot satisfying the above requirement (due to, for example, rapid merging of the SMBH binary in a time period shorter than the snapshot output frequency), the \child{} simulation is not included in the analysis.
We enforce a minimum of six \child{} simulations per family for that family to be analysed with our hierarchical modelling technique. 
For those \child{} simulations which pass our selection criteria, we take 30 random projections of the merger remnant, and determine the projected mass density in 50 radial bins evenly-spaced in logarithm space, with a minimum value of $0.2\,\mathrm{kpc}$ and a maximum value of $20.0\,\mathrm{kpc}$.
This allows us to build up a distribution of projected mass densities per radial bin, from which $\hat{\Sigma}$ and $\hat{\sigma}_\Sigma$ are determined.

\subsubsection{Latent Parameters for Observables}
With the omission of the parameter $\alpha$ in the modelling process, there are five latent parameters ($r_\mathrm{b}$, $R_\mathrm{e}$, $\tilde{\Sigma}_\mathrm{b}$, $\gamma$, and $n$, the set of which we denote as $\kappa$) that describe each \child{} simulation in a family. 
Here, we define $\tilde{\Sigma}_\mathrm{b} \equiv \Sigma_\mathrm{b}/10^{10}$, where $\Sigma_\mathrm{b}$ is defined in \autoref{eq:coresersic2}.
These latent parameters, depicted as non-hatted variables in blue circles in \autoref{fig:graphicalmodel}, are sampled to obtain a latent estimate on the projected mass density, $\Sigma$.
As each latent parameter in the set $\kappa$ is strictly non-negative, we use as prior distributions for each latent parameter a gamma distribution with hyperparameters $(\kappa_\alpha, \kappa_\beta)$, which are themselves drawn from a gamma distribution. 
To ensure our hyperparameter prior distributions are weakly-informative, the hyperparameters are drawn from the distributions in \autoref{tab:cshypers} for all simulation families. 
The sampled hyperparameter distributions are shown as corner plots in Figures \ref{fig:corner_params_1}-\ref{fig:corner_params_3}.

We ensure our hyperparameter and latent parameter distributions are representative of our observed data by performing prior predictive checks, an example of which is shown for the H-1.00 family in \autoref{fig:priorpred}. 
Critically, both the range of values spanned by the prior distributions and the general shape of the distribution suggests that the data is well represented by the model and the prior distributions on the hyper and latent parameters.

As likelihood function, we use the lognormal distribution $\mathrm{Lognorm}(\Sigma, \hat{\sigma}_\Sigma^2)$ as it a) has support on $[0,\infty)$, and b) produces normally-distributed variates in logarithm space, which is representative of the distribution of $\hat{\Sigma}$ for a given radial bin when we backwards-model and na\"ively bin our observations of $\hat{\Sigma}$ for that radial bin.

@@

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_posterior_pred_log10_proj_density_mean}.pdf}
    \caption{\drafting{Posterior predictive estimate for the core-S\'ersic fit for highest (fiducial) mass resolution. Darker shaded regions correspond to higher Bayesian Credible Intervals (BCIs). The observed data is overlaid, and seen to lie well within the 50\% BCI. }}
    \label{fig:postpred}
\end{figure*}

@@

\subsection{Latent Parameter Posterior Distributions}
We compute the posterior distribution using \textsc{Stan}, ensuring that the sampler exits without encountering any divergences, and that all \drafting{R-hat (is a summary stat, but how to draw? Don't want to confuse with observed radius)} values are within the range \drafting{$[0.99, 1.01]$}.
We first inspect how the posterior distribution appears visually by computing posterior predictive checks, an example of which is shown in \autoref{fig:postpred} for the family H-1.00. 
We find that the data is well described by the 50\% BCI \drafting{introduce what BCI is} of the posterior distribution, as desired.

As part of the hierarchical modelling sampling process, we obtain in addition to the posterior distribution on $\Sigma(R)$ the latent parameter distributions from which $\Sigma(R)$ is determined. 
We show an example of these latent parameter distributions in \autoref{fig:cs_compare_naive}, where the PDF of the latent parameter distributions is given in blue. 
For comparison, we also show the estimated uncertainty on each latent parameter determined using na\"ive Gaussian statistics.
To do this, we fit each density profile in logarithm space using a Least-Squares method, and then determine the median and IQR of the sample. 
The median value and spread for each latent parameter using this na\"ive method is shown above the corresponding posterior distribution estimate in \autoref{fig:cs_compare_naive}.
The Gaussian statistics fail to capture the correct estimates for the latent parameter estimates, with some parameters (e.g. $n$) converging to the boundary value we limit the Least Squares method to (for $n$, the bounds are $[0,20]$).
We try successively relaxing and tightening the boundary values of each parameter to obtain a `physically-realistic' value for each optimal latent parameter value, with no success. 
Henceforth, we abandon any attempt to use a Least Squares minimisation to recover the optimal latent parameter values, returning instead to our Bayesian approach.

@@

\begin{figure}
    \includegraphics{compare_binaries.pdf}
    \caption{\drafting{Show an example of the scatter we get from perturbations. These are plotted with t=0 corresponding to when the child run started. Best way to show this, or should a time-shift be done? This way at least we see clearly that different sims become bound at different times.}}
    \label{fig:ketju_scatter}
\end{figure}

\subsection{Binary Properties}
After considering the robustness of the projected density profile to SMBH Brownian motion, we turn our attention to a similar analysis of the orbital parameters of the SMBH binary, namely the semimajor axis $a$ and the eccentricity $e$. 
It is instructive to first consider the raw data of a particular simulation family, shown in \autoref{fig:ketju_scatter} for the case of H-1.00. 
Immediately apparent is the scatter in eccentricity between different realisations, with some runs (e.g. 006) having $e<0.2$ for the majority of the integration time, whilst other realisations (e.g. 002) have $e>0.9$ for the same time period.
As discussed in \drafting{section}, the efficiency of GW as a hardening mechanism is directly related to the binary eccentricity, and is highly non-linear. 
This is observed in those \child{} realisations that merged, and the timescale within which they merged.
As seen in \autoref{fig:ketju_scatter}, a higher eccentricity leads to more rapid BH binary mergers as opposed to a lower eccentricity. 
Out of the 50 simulations that we ran, we observe \drafting{XXX} systems that resulted in SMBH binary coalescence.

>>>>

@@

\begin{table}
    \centering
    \caption{The parameters for each hyperparameter distribution for the core-S\'ersic model, where each distribution is a normal distribution.}
    \label{tab:cshypers}
    \begin{tabular}{ccc}
        \hline
        Hyperparameter & $\mu$ & $\sigma$ \\
        \hline
        $r_{\mathrm{b}, \mu}$ & 0.0 & 2.0 \\
        $r_{\mathrm{b}, \sigma}$ & 0.0 & 1.0 \\
        $R_{\mathrm{e}, \mu}$ & 0.0 & 14.0 \\
        $R_{\mathrm{e}, \sigma}$ & 0.0 & 6.0 \\
        $\log_{10}\Sigma_{\mathrm{b}, \mu}$ & 0.0 & 20.0 \\
        $\log_{10}\Sigma_{\mathrm{b}, \sigma}$ & 0.0 & 10.0 \\
        $\gamma_\mu$ & 0.0 & 0.5 \\
        $\gamma_\sigma$ & 0.0 & 1.0 \\
        $n_\mu$ & 0.0 & 8.0 \\
        $n_\sigma$ & 0.0 & 5.0 \\
        $a_\mu$ & 0.0 & 20.0 \\
        $a_\sigma$ & 0.0 & 10.0 \\
        \hline
    \end{tabular}
\end{table}

@@

\subsubsection{Observable Quantities}
The hierarchical model takes three observable quantities, denoted by hat-variables in circles in \autoref{fig:graphicalmodel}. 
These variables are the radius $\hat{R}$, and the mean projected mass density $\hat{\Sigma}$, and the standard deviation of the projected mass density $\hat{\sigma}_\Sigma$ measured in the radial bins centred on $\hat{R}$.
For each \child{} simulation, we choose to `observe' the first snapshot where the binary semimajor axis is \textit{at most} $10\,\mathrm{pc}$. 
In the event that there is no snapshot satisfying the above requirement (due to, for example, rapid merging of the SMBH binary in a time period shorter than the snapshot output frequency), the \child{} simulation is not included in the analysis.
We enforce a minimum of six \child{} simulations per family for that family to be analysed with our hierarchical modelling technique. 
For those \child{} simulations which pass our selection criteria, we take 30 random projections of the merger remnant, and determine the projected mass density in 50 radial bins evenly-spaced in logarithm space, with a minimum value of $0.2\,\mathrm{kpc}$ and a maximum value of $20.0\,\mathrm{kpc}$.
This allows us to build up a distribution of projected mass densities per radial bin, from which $\hat{\Sigma}$ and $\hat{\sigma}_\Sigma$ are determined.

\subsubsection{Latent Parameters for Observables}
There are six latent parameters ($r_\mathrm{b}$, $R_\mathrm{e}$, $\log_{10}\Sigma_\mathrm{b}$, $\gamma$, $n$, and $\alpha$, the set of which we denote as $\kappa$) that describe each \child{} simulation in a family. 
These latent parameters, depicted as non-hatted variables in circles in \autoref{fig:graphicalmodel}, are sampled to obtain a latent estimate on the projected mass density, $\Sigma$.
The latent parameters in the set $\kappa$ are strictly non-negative, with the exception of $\log_{10}\Sigma_\mathrm{b}$. 
For these five positive-constrained latent parameters, we model the latent distribution as a half normal distribution. 
Each location hyperparameter for each of these latent parameters is also drawn from a truncated normal distribution, as a negative expectation value may be excluded on physical grounds.
For $\log_{10}\Sigma_\mathrm{b}$, we assume a normal distribution, with a location hyperparameter centred at $\log_{10}\Sigma_{\mathrm{b}, \mu}=10$. 
For all scale hyperparameters, we use a half normal distribution (as the scale of a distribution must of course always be positive).
To ensure our hyperparameter prior distributions are weakly-informative, the hyperparameters are drawn from the distributions in \autoref{tab:cshypers} for all simulation families. 
The sampled hyperparameter distributions are shown as corner plots in Figures \ref{fig:corner_params_1}-\ref{fig:corner_params_3}. \drafting{Corner plots maybe not so informative here, do we really expect correlations? Maybe 1D distributions better for the hyperpriors.}

We ensure our hyperparameter and latent parameter distributions are representative of our observed data by performing prior predictive checks, an example of which is shown for the H-1.00 family in \autoref{fig:priorpred}. 
Critically, both the range of values spanned by the prior distributions and the general shape of the distribution suggests that the data is well represented by the model and the prior distributions on the hyper and latent parameters. \drafting{Not yet, but will be. Some issue with the sampler.}

For each radial bin, we backwards-model and na\"ively bin our thirty observations of $\log_{10}\hat{\Sigma}$ for that radial bin to assess the likely distribution.
We find that the different projections of the mass density profile in a given radial bin are well-described by a normal distribution in logarithm space. 
Thus, as likelihood function, we use the normal distribution $\mathrm{Lognorm}(\Sigma, \hat{\sigma}_\Sigma^2)$.

@@

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_posterior_pred_log10_proj_density_mean}.pdf}
    \caption{\drafting{Posterior predictive estimate for the core-S\'ersic fit for highest (fiducial) mass resolution. Darker shaded regions correspond to greater Highest Density Intervals (HDIs). The observed data is overlaid, and seen to lie well within the 50\% HDIs. }}
    \label{fig:postpred}
\end{figure*}

@@

\subsection{Latent Parameter Posterior Distributions}
We compute the posterior distribution using \textsc{Stan}, ensuring that the sampler exits without encountering any divergences, and that all \drafting{R-hat (is a summary stat, but how to draw? Don't want to confuse with observed radius)} values are within the range \drafting{$[0.99, 1.01]$}.
We first inspect how the posterior distribution appears visually by computing posterior predictive checks, an example of which is shown in \autoref{fig:postpred} for the family H-1.00. 
We find that the data is well described by the 50\% HDI \drafting{introduce what HDI is} of the posterior distribution, as desired.

As part of the hierarchical modelling sampling process, we can obtain in addition to the posterior distribution on $\Sigma(R)$ the latent parameter distributions from which $\Sigma(R)$ is determined by marginalising out the other parameters. 
We show an example of these latent parameter distributions in \autoref{fig:cs_compare_naive}, where the PDF of the latent parameter distributions is given in blue. 
For comparison, we also show the estimated uncertainty on each latent parameter determined using na\"ive $\chi^2$ minimisation techniques.
To do this, we fit each density profile in logarithm space using a Least-Squares method, and then determine the median and IQR of the sample. 
The median value and spread for each latent parameter using this na\"ive method is shown above the corresponding posterior distribution estimate in \autoref{fig:cs_compare_naive}.
The $\chi^2$ minimisation fails to capture the correct estimates for the latent parameter estimates, with some parameters (e.g. $n$) converging to the boundary value we limit the Least Squares method to (for $n$, the bounds are $[0,20]$).
We try successively relaxing and tightening the boundary values of each parameter to obtain a `physically-realistic' value for each optimal latent parameter value, with no success. 
Henceforth, we abandon any attempt to use a Least Squares minimisation to recover the optimal latent parameter values, returning instead to our Bayesian approach.

@@

\subsection{Hardening Rate}
After considering the robustness of the projected density profile to SMBH Brownian motion, we turn our attention to a similar analysis of the hardening rate of the SMBH binary. 
The hardening rate $H$ in \autoref{eq:quinlan_H} connects the evolution of the inverse semimajor axis of the binary system to the surrounding stellar mass distribution. 
To extract the data for our hierarchical model for hardening rate, we take for each \child{} simulation the semimajor axis values between $a=a_\mathrm{h}$ and $a/\mathrm{pc}=10$.
We then fit the differential equation give in \autoref{eq:quinlan_H} to obtain a parameter estimate of the combined quantity $HG\rho/\sigma$.
We fit for the combined property in preference to just $H$ as the temporal-resolution of the semimajor axis $a$ is \drafting{check} $10^4$ times more frequent than the temporal resolution of the snapshots, requiring linear (or higher order) interpolation to be done between snapshots to obtain a continuous function of $G\rho/\sigma$ that can be mapped to the temporal resolution of $a$. 
To avoid these assumptions, we estimate the combined quantity $HG\rho/\sigma$.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{quinlan_posterior_pred_inv_a.png}
    \caption{\drafting{Example plot. This is an old one and needs to be updated, but gives us the idea of what we can talk about}}
    \label{fig:quinlan_H}
\end{figure}

\subsection{Binary Properties}
\begin{figure}
    \includegraphics{compare_binaries.pdf}
    \caption{\drafting{Show an example of the scatter we get from perturbations. These are plotted with t=0 corresponding to when the child run started. Best way to show this, or should a time-shift be done? This way at least we see clearly that different sims become bound at different times.}}
    \label{fig:ketju_scatter}
\end{figure}
\drafting{INtro here} orbital parameters of the SMBH binary, namely the semimajor axis $a$ and the eccentricity $e$. 
It is instructive to first consider the raw data of a particular simulation family, shown in \autoref{fig:ketju_scatter} for the case of H-1.00. 
Immediately apparent is the scatter in eccentricity between different realisations, with some runs (e.g. 006) having $e<0.2$ for the majority of the integration time, whilst other realisations (e.g. 002) have $e>0.9$ for the same time period.
As discussed in \drafting{section}, the efficiency of GW as a hardening mechanism is directly related to the binary eccentricity, and is highly non-linear. 
This is observed in those \child{} realisations that merged, and the timescale within which they merged.
As seen in \autoref{fig:ketju_scatter}, a higher eccentricity leads to more rapid BH binary mergers as opposed to a lower eccentricity. 
Out of the 50 simulations that we ran, we observe \drafting{XXX} systems that resulted in SMBH binary coalescence.

### 2023-03-01 papers/paper-1/sections/stats.tex
Start adding small section on hardening rate

Additionally, we must also consider the shape of each parameter's prior distribution, as parameters are generally constrained by some physical condition. 
For example, a normalising radius $a$ cannot take a value less than 0, so a normal distribution, which has an unbounded domain, is an inappropriate choice for the prior distribution of this variable.
Choosing a log-normal or gamma distribution with support on $[0,\infty)$ would represent a physically-motivated choice for the prior distribution here.

>>>>

Additionally, we must also consider the shape of each parameter's prior distribution, as parameters are generally constrained by some physical condition. 
For example, a normalising radius $a$ cannot take a value less than 0, so a normal distribution, which has an unbounded domain, is an inappropriate choice for the prior distribution of this variable.
Choosing a truncated normal or gamma distribution with support on $[0,\infty)$ would represent a physically-motivated choice for the prior distribution here.
